A quantum parallel Markov chain Monte Carlo

### 2021-11-10 main.tex
correct probabilities

Given a set of $N$ items and a function $f:\{0,1,\dots,N-1\}\rightarrow \{0,1\}$ that evaluates to $1$ for a single element, \citet{grover1996fast} develops an algorithm that uses quantum parallelism to score \emph{quadratic} speedups over its classical counterparts. After only $\order{\sqrt{N}}$ evaluations of $f(\cdot)$, Grover's algorithm returns the $x\in \{0,1,\dots,N-1\}$ satisfying $f(x) =1$ with probability at least $1/2$.  Compare this to $\order{N}$ requirement for the randomized classical algorithm that must evaluate $f(\cdot)$ over at least $N/2$ items to obtain the same probability of detecting $x$.  The algorithm takes the state $\ket{0}^{\otimes N}\ket{1}$ as input and applies Hadamard gates to each of the individual $N+1$ input qubits.  The resulting state is 
\begin{align*}
\ket{h}\ket{-}= \left(\frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \right)  \frac{1}{\sqrt{2}}\left(\ket{0} -\ket{1}\right) =  \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \ket{-} \, .
\end{align*}
Next, we apply the oracle gate $\U_f: \ket{x}\ket{y} \rightarrow \ket{x}\ket{y\oplus f(x)}$ and note that 
\begin{align*}
\U_f \ket{x} \ket{-} &= \U_f \ket{x} \frac{1}{\sqrt{2}}\left(\ket{0} -\ket{1}\right) 
= \frac{1}{\sqrt{2}} \big(\ket{x} \ket{0\oplus f(x)} - \ket{x}\ket{1\oplus f(x)} \big) \\
&= -1^{f(x)} \ket{x}\ket{-}  \, .
\end{align*}
Thus, $\U_f$ flips the sign for the state $\ket{x_0}$ for which $f(x_0)=1$ but leaves the other states unchanged.  If we suppress the ancillary qubit $\ket{-}$, then $\U_f$ is equivalent to the gate $\U_{x_0}$ defined as $\U_{x_0} \ket{x} = -1^{\delta_{x_0}}\ket{x}$. We may succinctly write this gate as the Householder matrix that reflects vectors about the unique hyperplane through the origin that has $\ket{x_0}$ for a normal vector:
\begin{align*}
\U_{x_0}= \I - 2 \ket{x_0}\bra{x_0} \, .
\end{align*}
The action of this gate on the state $\ket{h}$ takes the form
\begin{align*}
\frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \longrightarrow \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} -1^{\delta_{x_0}(x)} \ket{x} \, .
\end{align*}
Next, the algorithm reflects the current state about the hyperplane that has $\ket{h}$ as a normal vector and negates the resulting state:
\begin{align*}
\Big(2\ket{h}\bra{h}-\I \Big) \left( \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} -1^{\delta_{x_0}(x)} \ket{x}  \right)
&= \frac{1}{\sqrt{N}}\sum_{x} \left( (-1)^{1-\delta_{x_0}(x)} + 2 \frac{(N-1)}{N} \right)\ket{x}\\ &= \frac{(3N-2)}{N^{3/2}} \ket{x_0} + \sum_{x\neq x_0} \frac{(N-2)}{N^{3/2}}  \ket{x} \, .
\end{align*}
The scientist who measures the state at this moment would obtain the desired state $\ket{x_0}$ with slightly higher probability than the other individual states.

>>>>

Given a set of $N$ items and a function $f:\{0,1,\dots,N-1\}\rightarrow \{0,1\}$ that evaluates to $1$ for a single element, \citet{grover1996fast} develops an algorithm that uses quantum parallelism to score \emph{quadratic} speedups over its classical counterparts. After only $\order{\sqrt{N}}$ evaluations of $f(\cdot)$, Grover's algorithm returns the $x\in \{0,1,\dots,N-1\}$ satisfying $f(x) =1$ with probability at least $1/2$.  Compare this to $\order{N}$ requirement for the randomized classical algorithm that must evaluate $f(\cdot)$ over at least $N/2$ items to obtain the same probability of detecting $x$.  The algorithm takes the state $\ket{0}^{\otimes N}\ket{1}$ as input and applies Hadamard gates to each of the individual $N+1$ input qubits.  The resulting state is 
\begin{align*}
\ket{h}\ket{-}= \left(\frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \right)  \frac{1}{\sqrt{2}}\left(\ket{0} -\ket{1}\right) =  \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \ket{-} \, .
\end{align*}
Next, we apply the oracle gate $\U_f: \ket{x}\ket{y} \rightarrow \ket{x}\ket{y\oplus f(x)}$ and note that 
\begin{align*}
\U_f \ket{x} \ket{-} &= \U_f \ket{x} \frac{1}{\sqrt{2}}\left(\ket{0} -\ket{1}\right) 
= \frac{1}{\sqrt{2}} \big(\ket{x} \ket{0\oplus f(x)} - \ket{x}\ket{1\oplus f(x)} \big) \\
&= -1^{f(x)} \ket{x}\ket{-}  \, .
\end{align*}
Thus, $\U_f$ flips the sign for the state $\ket{x_0}$ for which $f(x_0)=1$ but leaves the other states unchanged.  If we suppress the ancillary qubit $\ket{-}$, then $\U_f$ is equivalent to the gate $\U_{x_0}$ defined as $\U_{x_0} \ket{x} = -1^{\delta_{x_0}}\ket{x}$. We may succinctly write this gate as the Householder matrix that reflects vectors about the unique hyperplane through the origin that has $\ket{x_0}$ for a normal vector:
\begin{align*}
\U_{x_0}= \I - 2 \ket{x_0}\bra{x_0} \, .
\end{align*}
The action of this gate on the state $\ket{h}$ takes the form
\begin{align*}
\frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \longrightarrow \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} -1^{\delta_{x_0}(x)} \ket{x} \, .
\end{align*}
Next, the algorithm reflects the current state about the hyperplane that has $\ket{h}$ as a normal vector and negates the resulting state:
\begin{align*}
\Big(2\ket{h}\bra{h}-\I \Big) \left( \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} -1^{\delta_{x_0}(x)} \ket{x}  \right)
&= \frac{1}{\sqrt{N}}\sum_{x} \left( (-1)^{1-\delta_{x_0}(x)} + 2 \frac{(N-2)}{N} \right)\ket{x}\\ &= \frac{(3N-4)}{N^{3/2}} \ket{x_0} + \sum_{x\neq x_0} \frac{(N-4)}{N^{3/2}}  \ket{x} \, .
\end{align*}
The scientist who measures the state at this moment would obtain the desired state $\ket{x_0}$ with a slightly higher probability of $(3N-4)^2/N^3$  than the individual probabilities of $(N-4)^2/N^3$ for the other states.
